CS378: Final Project - Data Artifacts \\ \url{https://github.com/sarimaleem/cs378_fp}

### 2023-04-27 paper/main.tex
final paper

This paper describes the training and testing of an ELECTRA model for natural
language inference using the Stanford Natural Language Inference Dataset. The
model's performance is benchmarked using contrast sets, and data cartography is
used to improve the model's accuracy. The results show that the base ELECTRA
model performs well on the dataset, and extracting ambiguous examples from data
cartography and training the model on them is an effective method for improving
its performance for some aspects.

@@

In order to reduce ambiguity between words, we decided to train the model on
ambiguous data - data with high variability and medium confidence. We found that
training that although training the model on ambiguous data reduced the models
overall performance, the performance of the model increased on adversarial data,
which we dubbed contrast sets.

@@

The model that was used was the ELECTRA model \cite{clark2020electra}. We are
training  on the stanford NLI dataset \cite{bowman2015large}. The dataset has
570000 training examples and 10000 validation examples. The baseline ELECTRA
modeled was fine tuned to the SNLI training set and was trained with three
epochs. We used the default loss function from the hugging face transformers
library.

For the fix, we took 1655 ambiguous data examples from the first 50000 training
examples and trained on those for 15 epochs on top of the original model. We
also trained 6915 difficult to learn examples for 5 epochs on top of the
baseline model and evaluate that.

@@

\subsection{Confusion Matrix Analysis}
In general, the model seems to do well with Separating Entailment and
Contradiction. For example, it only incorrectly identifies 2.47\% of entailment
examples as contradiction and 2.37\% of contradiction examples as entailement.

However, the model struggles more to understand the relationship between,
entailment and neutral, as well as contradiction and neutral. For example, it
mispredicted over 7\% of entailments and contradictions as neutrals. As long as 
mispredicting over 6\% of neutrals as entailments and over 7\% of of neutrals as
contradictions.

@@

In the example with hypothesis \textit{A man and a woman are looking at produce
on display} and premise \textit{A man and woman are staring at heads of
lettuce}. The model is unable to distinguish that lettuce is a type of produce
and instead makes assumption that they are different, thus leading it to think
they are contradictions.

@@

In their article \textit{Evaluating models' local decision boundaries via
contrast sets} \cite{gardner2020evaluating}, Gardner et al. explain the concept
of a \textit{contrast set}. A contrast set is a small but meaninful alteration
in the input data that typically leads to a different gold label.

In order to test if the model is finding artifacts in the data or genuinely
evaluating it, we annotated 35 different examples in the validation data that
the model predicted correctly, and then tested to see if the model was able to
predict them correctly again.

The results of the model showed a significant dip in accuracy, with accuracy
going down to 77\% from 89\%. The model mispredicted 8/35 examples. One pattern
that emerged from the mispredictions is that the model fails to see correlations
between various words and also fails to understand the relationships between
various words. The following is an example of this phenomenon: \\

@@

It is clear that the issue with the model is that it fails to understand
ambiguous training examples. This is especially seen given that the model
largely confuses, as seen earlier, neutral and entailment and instances of
contradiction.

@@

In order to deal with this, we first created a program that finds ambiguous
examples. We found the variance of the training example based on its confidence
of the gold label from a modeled trained on 500 iterations, 1000 iterations,
1500 iterations, 2000 iterations, 2500 iterations, and 3000 iterations. If th
variance was greater than 0.15, and the confidence of the model was in between
0.4 and 0.8, the example was deemed ambiguous. According to \textit{Swayamdipta
et al.}, training solely ambiguous examples leads to to greater performance
than training on the whole dataset, only easy to learn examples, or only hard to
learn examples. Thus it is logical to train on these examples.

@@

@@

However, the model that was trained on ambiguous data seemed to suffer on the
overall benchmark of the evaluation set. The accuracy dropped from 88\% to 84\%
on the validation set.

The most surprising results were the difficult to learn examples. The accuracy
of that model was significantly lower than base model, clocking in an accuracy
of 13\%. Even a model with random guessing could have done better.

@@

In this paper, we evaluate an ELECTRA based ML model on the Stanford Natural
Language Inference dataset. We find that the model has difficulty
differentiating between neutral labels and other classes and that it finds at
times spurious correlations between the premise and hypothesis. We also find
that the model suffers significant performance degradation when tested against a
contrast set. To remedy this, we recommend finding ambiguous labels (labels with
high variablity and medium conference) using data cartography, and train the
model on those data points in addition to the training set. Although not a
perfect method, training on ambiguous labels can help the model learn more
difficult correlations.

>>>>

This paper describes the training and testing of an ELECTRA model for natural
language inference using the Stanford Natural Language Inference Dataset. The
model's performance is benchmarked using contrast sets, and data cartography is
used to improve the model's accuracy. The results show that the base ELECTRA
model performs well on the dataset, and extracting \enquote{ambiguous} examples
from data cartography and training the model on them is an effective method for
improving its performance for some aspects.  We also found that
\enquote{difficult to learn} training examples vastly reduce model performance,
and suggest removing not having them in the training data.

@@

In order to reduce ambiguity between words, we decided to train the model on
ambiguous data - data with high variability and medium confidence. We found that
training that although training the model on ambiguous data reduced the models
overall performance by about 3\%, the performance of the model increased on
adversarial data, which in this case are contrast sets, by 3\% as well. We also
found that \enquote{difficult to learn} examples significantly lowered accuracy,
and belive model performance can be improved with them removed.

@@

The model that was used was the ELECTRA model \cite{clark2020electra}, a
transformer based on BERT. We are training  on the stanford NLI dataset
\cite{bowman2015large}. The dataset has 570000 training examples and 10000
validation examples. The baseline ELECTRA model was fine tuned to the SNLI
training set and was trained with three epochs. We used the default loss
function from the hugging face transformers library.

For the fix, we took 1655 ambiguous data examples from the first 50000 training
examples and trained on those for 15 epochs on top of the original model. We
also trained 6915 difficult to learn examples for 5 epochs on top of the
baseline model and evaluate that. Our evaluation was accuracy for each of our
different test/validation sets as well as a confusion matrix that describes the
distribution of the data.

@@

\subsection{Confusion Matrix Analysis}

In general, the model seems to do well with Separating Entailment and
Contradiction. As can be seen in Table 2, it only incorrectly identifies 2.47\%
of entailment examples as contradiction and 2.37\% of contradiction examples as
entailement.

However, the model struggles more to understand the relationship between,
entailment and neutral, as well as contradiction and neutral. For example, it
mispredicted over 7\% of entailments and contradictions as neutrals.
Furthermore, it also mispredicts over 6\% of neutrals as entailments and over
7\% of of neutrals as contradictions.

@@

In the example with hypothesis \textit{A man and a woman are looking at produce
on display} and premise \textit{A man and woman are staring at heads of
lettuce}. The model is unable to distinguish that lettuce is a type of produce
and instead makes assumption that they are different, thus leading it to think
they are contradictions. Further examples of this phenomenon can be seen in
Table 4, which gives examples.

@@

In their article \textit{Evaluating models' local decision boundaries via
contrast sets} \cite{gardner2020evaluating}, Gardner et al. explain the concept
of a \textit{contrast set}. A contrast set is a small but meaningful alteration
in the input data that typically leads to a different gold label.

In order to test if the model is finding artifacts in the data or genuinely
evaluating it, we hand annotated 35 different examples in the validation data that
the model predicted correctly, and then tested to see if the model was able to
predict them correctly again.

The results of the model showed a significant dip in accuracy, with accuracy
going down to 77\% from initially predicting them all correctly. Not to mention,
the statistic is significantly lower than the model baseline of 88\% The model
mispredicted 8/35 examples. One pattern that emerged from the mispredictions is
that the model fails to see correlations between various words and also fails to
understand the relationships between various words. The following is an example
of this phenomenon: \\

@@

It is clear that the issue with the model is that it fails to understand
ambiguous training examples. This is especially seen given that the model
largely confuses the semantics of various words that otherwise have similar
meanings.

@@

We believe that dataset cartography is a good fix to the problem. This is
because ambiguous training examples typically map to \enquote{hard} problems,
where as \enquote{hard to learn examples} typically map to training data with
errors. Therefore, we recommend training without \enquote{hard to learn}
examples and weighting \enquote{ambiguous} examples greater.

In order to deal with this, we first created a program that finds ambiguous
examples. We found the variance of the training example based on its confidence
of the gold label from a modeled trained on 500 iterations, 1000 iterations,
1500 iterations, 2000 iterations, 2500 iterations, and 3000 iterations. If the
variance was greater than 0.15, and the confidence of the model was in between
0.4 and 0.8, the example was deemed ambiguous. We further deemed labels with
confidence below 0.5 as \enquote{hard to learn} According to \textit{Swayamdipta
et al.}, training solely ambiguous examples leads to to greater performance than
training on the whole dataset, only easy to learn examples, or only hard to
learn examples. Thus it is logical to train on these examples.

@@

\subsection{Ambiguous Examples}

@@

However, the model that was trained on ambiguous data seemed to suffer on the
overall benchmark of the evaluation set. The accuracy dropped from 88\% to 84\%
on the validation set. Table 5 and 6 show the accuracy of the ambiguously
trained model on the contrast and validation set, and shows that there is
improvement in contrast set but some regression in the validation set.

\subsection{Difficult to learn examples}

Unsurprisingly, difficult to learn examples had accuracies that were
significantly lower than the base model. However, it was surpring the extent to
which they degraded the performance of the model. Like the ambiguous examples,
the difficult to learn examples were trained directly on top of the base model,
so the model was already partially fine tuned. Nevertheless, the 6915 training
examples over 5 epochs degraded the model performance a very large amount. The
accuracy of that model was significantly lower than base model, clocking in an
accuracy of 13\%. Even a model with random guessing could have done better.
Because of this, we highly suggest removing difficult to learn examples from the
training data.

@@

In this paper, we evaluate an ELECTRA based ML model on the Stanford Natural
Language Inference dataset. We find that the model has difficulty
differentiating between neutral labels and other classes and that it finds at
times spurious correlations between the premise and hypothesis. We also find
that the model suffers significant performance degradation when tested against a
contrast set. To remedy this, we recommend finding ambiguous labels (labels with
high variablity and medium conference) using data cartography, and train the
model on those data points in addition to the training set. Although not a
perfect method, training on ambiguous labels can help the model learn more
difficult correlations. We also suggest removing \enquote{difficult to learn}
training examples, as they hurt model performance and are likely annotator
errors. Dataset cartography overall is a highly effective method eliminating
unwanted training examples or weight wanted training examples.
